Program Transformations for miniKanren

### 2020-07-07 papers/aplas/text/evaluation.tex
Edits

The conservative partial deduction first unfolds those calls which are selected according to the heuristics.
Since exploring the implementations of boolean connectives makes more sense, they are unfolded before recursive calls of \lstinline{eval$^o$}.
The way conservative partial deduction treats this program is the same as it treats the other implementation in which boolean connectives
are moved to the left, as shown in Listing~\ref{eval:fst}.
This program is easier for \ecce to specialize which demonstrates how unequal is the behaviour of CPD for similar programs.

@@

Most of the redundancy of the generated program is removed by specialization with respect to the known type of the term.
This is why both implementations have comparable speed after specialization.
\ecce shows bigger speedup for the hand-written than ConsPD and vice versa for the generated implementation.
We believe that this difference can be explained with too much unfolding.
\ecce performs a lot of excessive unfolding for the generated program and only barely changes the hand-written program.
At the same time ConsPD specializes both implementations to comparable programs performing average amount of unfolding.
This shows that the heuristics we presented gives more stable, although not the best, results.

@@

Besides evaluator of logic formulas we also run the specializers on the relation \lstinline{unify$^o$} which searches for a unifier of two terms and a relation \lstinline{isPath$^o$}
specialized to search for paths in the graph.
Both relations were generated from functional programs by unnesting.
These two relations are described in paper~\cite{lozov2019relational} so we will not go into too many details here.

>>>>

The conservative partial deduction first unfolds those calls which are selected according to the heuristics.
Since exploring the implementations of boolean connectives makes more sense, they are unfolded before recursive calls of \lstinline{eval$^o$}.
The way conservative partial deduction treats this program is the same as it treats the other implementation in which boolean connectives
are moved to the left, as shown in Listing~\ref{eval:fst}.
This program is easier for \ecce to specialize which demonstrates how unequal the behaviour of CPD for similar programs is.

@@

Most of the redundancy of the generated program is removed by specialization with respect to the known type of the term.
This is why both implementations have comparable speed after specialization.
\ecce shows bigger speedup for the hand-written than ConsPD and vice versa for the generated implementation.
We believe that this difference can be explained by too much unfolding.
\ecce performs a lot of excessive unfolding for the generated program and only barely changes the hand-written program.
At the same time ConsPD specializes both implementations to comparable programs performing average amount of unfolding.
This shows that the heuristics we presented gives more stable, although not the best, results.

@@

Besides evaluator of logic formulas we also run the specializers on the relation \lstinline{unify$^o$}, which searches for a unifier of two terms, and a relation \lstinline{isPath$^o$},
used to search for paths in the graph.
Both relations were generated from functional programs by unnesting.
These two relations are described in paper~\cite{lozov2019relational} so we will not go into too many details here.

### 2020-07-07 papers/aplas/text/intro.tex
Edits

In this paper, we show on examples some issues which conjunctive partial deduction faces.
We also describe a novel approach to partial deduction of a relational programming language \mk.
We compare it to the existing specialization system (\ecce) on several programs and discuss why some \mk programs run slower after specialization.

>>>>

In this paper, we show on examples some issues which conjunctive partial deduction faces.
We also describe a novel approach to partial deduction of a relational programming language \mk.
We compare the described approach to the existing specialization system (\ecce) on several programs and discuss why some \mk programs run slower after specialization.

### 2020-07-07 papers/aplas/text/nonconj.tex
Conclusion rewritten
Edits

The algorithm pseudocode is shown in fig.~\ref{fig:ncpd-pseudo}.
For the sake of brevity and clarity, we provide functions \code{drive\_disj} and \code{drive\_conj} which describe how to process disjunctions and conjunctions respectively.
Driving itself is a trivial combination of presented functions (line 2).

A driving process creates a process tree, from which a residual program is later created.
The process tree is meant to mimic the execution of the input program.
The nodes of the process tree include a \emph{configuration} which describes the state of program evaluation at some point.
In our case configuration is a conjunction of relation calls.
The substitution computed at each step is also stored in the tree node, although it is not included in the configuration.

Hereafter, we consider all goals and relation bodies to be in \emph{canonical normal form}~---~a disjunction of conjunctions of either calls or unifications.
Moreover, we assume all fresh variables to be introduced into the scope and all unifications to be computed at each step.
Those disjuncts in which unifications fail are removed.
Other disjuncts take the form of possibly empty conjunction of relation calls accompanied with a substitution computed from unifications.
Any \mk term can be trivially transformed into the described form.
The function \code{normalize} in fig.~\ref{fig:ncpd-pseudo}  is assumed to perform term normalization.
The code is omitted for brevity.

@@

This process creates branchings whenever a disjunction is examined (lines 4--5).
At each step, we make sure that we do not start driving conjunction which we have already examined.
To do this, we check if the current conjunction is a renaming of any other configuration in the tree (line 11).
If it is, then we fold the tree by creating a special node which then is residualized into a call to the corresponding relation.

We decided not to perform generalization in this approach in the same fashion as CPD or supercompilation does.
Our conjunctions are always split into individual calls and are joined back together only if it is sensible.
If the need for generalization arises, i.e. homeomorphic embedding of conjunctions~\cite{de1999conjunctive} is detected, then we immediately stop driving this conjunction (line 12).
When residualizing such conjunction, we just generate a conjunction of calls to the input program before specialization.

@@

Unfolding in our case is done by substitution of some relation call by its body with simultaneous normalization and computation of unifications.
% To unfold a relation call we do the following steps.
% \db{TODO: the following has alredy mentioned above.}
% First, the formal arguments of a relation are substituted for the actual arguments of the call in the body.
% All fresh variables get instantiated.
% The body is transformed into a canonical form (disjunction of conjunctions of either calls or unifications).
% All unifications are computed.
% Those disjuncts in which unifications fails are removed.
% Other disjuncts take form of a conjunction of relation calls accompanied with a substitution.
The unfolding itself is straightforward however it is not always clear what to unfold and when to \emph{stop} unfolding.
Unfolding in functional programming languages specialization, as well as inlining in imperative one, is usually considered to be safe from the residual program efficiency point of view.
It may only lead to code explosion or code duplication which is mostly left to a target program compiler optimization or even is out of consideration at all if a specializer is considered as a standalone tool~\cite{jonesbook}.

Unfortunately, this is not the case for the specialization of relational programming language.
Unlike functional and imperative, in logic and relational programming language unfolding may easily affect the target program efficiency~\cite{leuschel2002logic}.
Unfolding too much may create extra unifications, which is by itself a costly operation, or even introduce duplicated computations by propagating the unfolding results onto neighbouring conjuncts.

@@

This heuristics is aimed at selecting the relation call within a conjunction which is both safe to unfold and may lead to discovering contradictions within the conjunction.
The unsafe unfolding leads to an uncontrollable increase in the number of relation calls in a conjunction.
It is best to first unfold those relation calls which can be fully computed up to substitutions.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\textwidth]{heuristic-crop.pdf}
  \caption{Heuristic selection pseudocode}
  \label{fig:heu-pseudo}
\end{figure}

@@

The pseudocode describing heuristics is shown in fig.~\ref{fig:heu-pseudo}.
Selecting a good relation call can fail (line 1).
The implementation works such that we first select those relation calls which are static, and only if there are none, we proceed to consider deterministic unfoldings and then we search for those which are less branching.
We believe this heuristics provides a good balance in unfolding.
% The final heuristic selects the first conjunct which suites either of the following cases.
% First we unfold those conjuncts which are static.
% Then --- deterministic.
% Then those which are less branching.
% The last to be unfolded are those calls, which unfold to a substitution with not conjunction.

>>>>

The algorithm pseudocode is shown in Fig.~\ref{fig:ncpd-pseudo}.
For the sake of brevity and clarity, we provide functions \code{drive\_disj} and \code{drive\_conj} which describe how to process disjunctions and conjunctions respectively.
Driving itself is a trivial combination of the functions provided (line 2).

A driving process creates a process tree, from which a residual program is later created.
The process tree is meant to mimic the execution of the input program.
The nodes of the process tree include a \emph{configuration} which describes the state of program evaluation at some point.
In our case a configuration is a conjunction of relation calls.
The substitution computed at each step is also stored in the tree node, although it is not included in the configuration.

Hereafter, we consider all goals and relation bodies to be in \emph{canonical normal form}~---~a disjunction of conjunctions of either calls or unifications.
Moreover, we assume all fresh variables to be introduced into the scope and all unifications to be computed at each step.
Those disjuncts in which unifications fail are removed.
Each other disjunct takes the form of a possibly empty conjunction of relation calls accompanied with a substitution computed from unifications.
Any \mk term can be trivially transformed into the described form.
The function \code{normalize} in Fig.~\ref{fig:ncpd-pseudo} is assumed to perform term normalization.
The code is omitted for brevity.

@@

This process creates branchings whenever a disjunction is examined (lines 4--5).
At each step, we make sure that we do not start driving a conjunction which we have already examined.
To do this, we check if the current conjunction is a renaming of any other configuration in the tree (line 11).
If it is, then we fold the tree by creating a special node which then is residualized into a call to the corresponding relation.

In this approach, we decided not to generalize in the same fashion as CPD or supercompilation.
Our conjunctions are always split into individual calls and are joined back together only if it is sensible.
If the need for generalization arises, i.e. homeomorphic embedding of conjunctions~\cite{de1999conjunctive} is detected, then we immediately stop driving this conjunction (line 12).
When residualizing such a conjunction, we just generate a conjunction of calls to the input program before specialization.

@@

Unfolding in our case is done by substitution of some relation call by its body with simultaneous normalization and computation of unifications.
% To unfold a relation call we do the following steps.
% \db{TODO: the following has alredy mentioned above.}
% First, the formal arguments of a relation are substituted for the actual arguments of the call in the body.
% All fresh variables get instantiated.
% The body is transformed into a canonical form (disjunction of conjunctions of either calls or unifications).
% All unifications are computed.
% Those disjuncts in which unifications fails are removed.
% Other disjuncts take form of a conjunction of relation calls accompanied with a substitution.
The unfolding itself is straightforward however it is not always clear what to unfold and when to \emph{stop} unfolding.
Unfolding in the context of specialization of functional programming languages, as well as inlining in specialization of imperative languages, is usually considered to be safe from the residual program efficiency point of view.
It may only lead to code explosion or code duplication which is mostly left to a target program compiler optimization or even is out of consideration at all if a specializer is considered as a standalone tool~\cite{jonesbook}.

Unfortunately, this is not the case for the specialization of a relational programming language.
Unlike in functional and imperative languages, in logic and relational programming languages unfolding may easily affect the target program's efficiency~\cite{leuschel2002logic}.
Unfolding too much may create extra unifications, which is by itself a costly operation, or even introduce duplicated computations by propagating the results of unfolding onto neighbouring conjuncts.

@@

This heuristics is aimed at selecting a relation call within a conjunction which is both safe to unfold and may lead to discovering contradictions within the conjunction.
An unsafe unfolding leads to an uncontrollable increase in the number of relation calls in a conjunction.
It is best to first unfold those relation calls which can be fully computed up to substitutions.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\textwidth]{heuristic-crop.pdf}
  \caption{Heuristics selection pseudocode}
  \label{fig:heu-pseudo}
\end{figure}

@@

The pseudocode describing our heuristics is shown in Fig.~\ref{fig:heu-pseudo}.
Selecting a good relation call can fail (line 1).
The implementation works such that we first select those relation calls which are static, and only if there are none, we proceed to consider deterministic unfoldings and then we search for those which are less branching.
We believe this heuristics provides a good balance in unfolding.
% The final heuristic selects the first conjunct which suites either of the following cases.
% First we unfold those conjuncts which are static.
% Then --- deterministic.
% Then those which are less branching.
% The last to be unfolded are those calls, which unfold to a substitution with not conjunction.

### 2020-07-07 papers/aplas/text/related.tex
Edits

Specialization is an attractive technique aimed to improve the performance of a program if some of its arguments are known statically.
It is studied for functional, imperative and logic programing and comes in different forms: partial evaluation~\cite{jonesbook} and partial deduction~\cite{lloyd1991partial}, supercompilation~\cite{soerensen1996positive}, distillation~\cite{hamilton2007distillation} and many more.

The heart of supercompilation-based techniques is \emph{driving}~---~a symbolic execution of a program through all possible execution paths.
The result of driving is so-called \emph{process tree} where nodes correspond to \emph{configurations} which present computation state, for example, a term in case of pure functional programming languages.
Each path in the tree corresponds to some concrete program execution.
The two main sources for supercompilation optimizations are aggressive information propagation about variable values, equalities, and disequalities and precomputing of all deterministic semantic evaluation steps, i.e. combining of consecutive process tree nodes with no branching, also known as \emph{deforestation}~\cite{deforestation}.
When the tree is constructed, the resulting, or \emph{residual}, program can be extracted from the process tree by the process called \emph{residualization}.
Of course, process tree can contain infinite branches.
\emph{Whistles} --- heuristics to identify possibly infinite branches --- are used to ensure termination in supercompilation.
If the whistle signalled during the construction of some branch, then something should be done to ensure termination.
The most common approaches include either stopping driving the infinite branch completely (no specialization is done in this case and the source code is blindly copied into the residual program) or folding the process tree to a \emph{process graph}.
The main instrument to perform such a folding is \emph{generalization}, i.e. abstracting away some computed data about the current term which makes folding possible.
One source of infinite branches is consecutive recursive calls to the same function with an accumulating parameter: by unfolding such a call further one can only increase the term size which leads to nontermination.
The accumulating parameter can be removed by replacing the call with its generalization.
There are several ways to ensure process correctness and termination, the most common being \emph{homeomorphic embedding}~\cite{Higman52,Kruskal60} used as a whistle and most-specific generalization of terms.
% When two dangerously similar nodes are encountered,
% For example, two consecutive recursive calls of a function with accumulating parameter, and the first is not an instance of the second, then one may construct a new, generalized, node such that both of original nodes are instances of the last one, then one of the initial nodes is replaced with the generalized one.

While supercompilation generally improves the behaviour of input programs and distillation can even provide superlinear speedup, there are no ways to predict the effect of specialization on a given program in the general.
What is worse, they rarely consider the residual program efficiency from the target language evaluator point of view.
The main optimization source is computing in advance all possible intermediate and statically-known semantics steps at program transformation-time.
Other criteria, like the size of the generated program or possible optimizations and execution cost of different language constructions by the target language evaluator, are usually out of consideration~\cite{jonesbook}.
It is known that supercompilation may adversely affect GHC optimizations yielding standalone compilation more powerful~\cite{SCBE,TCES} and cause code explosion~\cite{SCHC}.
Moreover, it may be hard to predict the real speedup of any given program on concrete examples even disregarding problems above because of the complexity of the transformation algorithm.
The worst-case for partial evaluation is when all static variables are used in a dynamic context, and there is some advice on how to implement a partial evaluator as well as a target program so that specialization indeed improved its performance~\cite{jonesbook,bulyonkov84}.
There is lack of research in determining the classes of programs which transformers would definitely speed up.

Conjunctive partial deduction~\cite{de1999conjunctive} makes an effort to provide reasonable control for left-to-right evaluation strategy of \pro.
CPD constructs a tree which models goal evaluation and is similar to SLDNF tree, then a residual program is generated from this tree.
Partial deduction itself resembles driving in supercompilation~\cite{gluck1994partial}.
The specialization is done in two levels of control: the local control determines the shape of the residual programs, while the global control ensures that every relation which can be called in the residual program is indeed defined.
The leaves of local control trees become nodes of the global control tree.
CPD analyses these nodes at the global level and runs local control for all those which are new.

@@

An empitical study shown that the most well-behaved strategy of local control in CPD for \pro is \emph{deterministic unfolding}~\cite{leuschel1997advanced}.
An atom is unfolded only if only one suitable clause head exists for it with the one exception: it is allowed to unfold an atom non-deterministically once for one local control tree.
This means that if a non-deterministic atom is the leftmost within conjunction, it is most likely to be unfolded and introduce many new relation calls within the conjunction.
We believe this is the core problem with CPD which limits its power when applied to \mk.
The strategy of unfolding atoms from left to right is reasonable in the context of \pro because it mimics the way programs in \pro execute.
But it often leads to larger global control trees and, as a result, bigger, less efficient programs.
The evaluation result of a \mk program does not depend on the order of atoms (relation calls) within a conjunction, thus we believe a better result can be achieved by selecting a relation call which can restrict the number of branches in the tree.
We describe our approach which implements this idea in the next section.

>>>>

Specialization is an attractive technique aimed to improve the performance of a program if some of its arguments are known statically.
Specialization is studied for functional, imperative and logic programing and comes in different forms: partial evaluation~\cite{jonesbook} and partial deduction~\cite{lloyd1991partial}, supercompilation~\cite{soerensen1996positive}, distillation~\cite{hamilton2007distillation} and many more.

The heart of supercompilation-based techniques is \emph{driving}~---~a symbolic execution of a program through all possible execution paths.
The result of driving is so-called \emph{process tree} where nodes correspond to \emph{configurations} which represent computation state.
For example, in the case of pure functional programming languages, the computational state might be a term.
Each path in the tree corresponds to some concrete program execution.
The two main sources for supercompilation optimizations are aggressive information propagation about variables' values, equalities and disequalities, and precomputing of all deterministic semantic evaluation steps.
The latter process, also known as \emph{deforestation}~\cite{deforestation}, means  combining of consecutive process tree nodes with no branching.
When the tree is constructed, the resulting, or \emph{residual}, program can be extracted from the process tree by the process called \emph{residualization}.
Of course, process tree can contain infinite branches.
\emph{Whistles} --- heuristics to identify possibly infinite branches --- are used to ensure supercompilation terminates.
If a whistle signalls during the construction of some branch, then something should be done to ensure termination.
The most common approaches include stop driving the infinite branch completely (no specialization is done in this case and the source code is blindly copied into the residual program) or fold the process tree to a \emph{process graph}.
The main instrument to perform such a folding is \emph{generalization}.
Generalization, abstracting away some computed data about the current term, makes folding possible.
%  i.e. abstracting away some computed data about the current term which makes folding possible.
One source of infinite branches is consecutive recursive calls to the same function with an accumulating parameter: by unfolding such a call further one can only increase the term size which leads to nontermination.
The accumulating parameter can be removed by replacing the call with its generalization.
There are several ways to ensure process correctness and termination, most-specific generalization
(anti-unification) and using \emph{homeomorphic embedding}~\cite{Higman52,Kruskal60} as a
whistle being the most common.
%the most common being \emph{homeomorphic embedding}~\cite{Higman52,Kruskal60} used as a whistle and most-specific generalization of terms.
% When two dangerously similar nodes are encountered,
% For example, two consecutive recursive calls of a function with accumulating parameter, and the first is not an instance of the second, then one may construct a new, generalized, node such that both of original nodes are instances of the last one, then one of the initial nodes is replaced with the generalized one.

While supercompilation generally improves the behaviour of input programs and distillation can even provide superlinear speedup, there are no ways to predict the effect of specialization on a given program in the general.
What is worse, the efficiency of residual program from the target language evaluator point of view is rarely considered in the literature.
The main optimization source is computing in advance all possible intermediate and statically-known semantics steps at program transformation-time.
Other criteria, like the size of the generated program or possible optimizations and execution cost of different language constructions by the target language evaluator, are usually out of consideration~\cite{jonesbook}.
It is known that supercompilation may adversely affect GHC optimizations yielding standalone compilation more powerful~\cite{SCBE,TCES} and cause code explosion~\cite{SCHC}.
Moreover, it may be hard to predict the real speedup of any given program on concrete examples even disregarding the problems above because of the complexity of the transformation algorithm.
The worst-case for partial evaluation is when all static variables are used in a dynamic context, and there is some advice on how to implement a partial evaluator as well as a target program so that specialization indeed improves its performance~\cite{jonesbook,bulyonkov84}.
There is a lack of research in determining the classes of programs which transformers would definitely speed up.

Conjunctive partial deduction~\cite{de1999conjunctive} makes an effort to provide reasonable control for the left-to-right evaluation strategy of \pro.
CPD constructs a tree which models goal evaluation and is similar to a SLDNF tree, then a residual program is generated from this tree.
Partial deduction itself resembles driving in supercompilation~\cite{gluck1994partial}.
The specialization is done in two levels of control: the local control determines the shape of the residual programs, while the global control ensures that every relation which can be called in the residual program is indeed defined.
The leaves of local control trees become nodes of the global control tree.
CPD analyses these nodes at the global level and runs local control for all those which are new.

@@

An empitical study shown that the most well-behaved strategy of local control in CPD for \pro is \emph{deterministic unfolding}~\cite{leuschel1997advanced}.
An atom is unfolded only if precisely one suitable clause head exists for it with the one exception: it is allowed to unfold an atom non-deterministically once for one local control tree.
This means that if a non-deterministic atom is the leftmost within a conjunction, it is most likely to be unfolded and to introduce many new relation calls within the conjunction.
We believe this is the core problem with CPD which limits its power when applied to \mk.
The strategy of unfolding atoms from left to right is reasonable in the context of \pro because it mimics the way programs in \pro execute.
But it often leads to larger global control trees and, as a result, bigger, less efficient programs.
The evaluation result of a \mk program does not depend on the order of atoms (relation calls) within a conjunction, thus we believe a better result can be achieved by selecting a relation call which can restrict the number of branches in the tree.
We describe our approach which implements this idea in the next section.
